\chapter{Assembly algorithms}

There are several formulations of the genome assembly problem, but
most of them are not user in practice.

First of them is a variation of shortest common superstring problem, where we
almost account for the possibility of sequencing errors \cite{kececioglu1995combinatorial}. 

\begin{definition}{DNA sequence reconstruction problem 1.}
Given set of reads $\mathcal{F}$ and error rate $\varepsilon$, find
a shortest sequence $S$ such that for every $A \in \mathcal{F}$ there is a substring
$B$ of $S$ such that:
$$min(d(A,B), d(\bar{A}, B)) \leq \varepsilon |A|$$
\end{definition}

This formulation has several problems. 
Firstly, it compresses repeated regions in genome. For example take sequence
$ACGGGGGGGGGGGGTT$ and all of its substrings of length 3. Then the correct
output for our formulation is $ACGGGTT$.

Secondly, it is ambiguous. For example take sequence
$ACGGGGTATGGGGCTCGGGGAA$ and all of its substring of length 3. Then the correct
outputs for our formulation are $ACGGGGTATGGGGCTCGGGGAA$ and
$ACGGGGCTCGGGGTATGGGGAA$.

A better formulation is given by \cite{myers1995toward}. It also considers
coverage of output sequence by substring and wants it to be as uniform as possible.

We consider our reconstructed string $S$ and the layout consisting of
$F$ pairs of integers $(s_i, e_i)$, which indicates starting and ending positions
of reads in the reconstructed sequence. The layout is $\varepsilon$-valid
if for each read $A$ the edit distance between $S[s_i:e_i]$ and the read
is at most $\varepsilon |A|$.

We will now formalize the notion of uniform coverage. Lets consider a observed
distribution of read start points (the proportion of reads which start before $x$):
$$D_{obs}(x) = \frac{|\{s_i < x\}|}{F}$$

We now consider a source distribution of a sampling process $D_{src}$ (which is usually
uniform, but can be nonuniform due to some systematic errors) and define maximum
deviation between these two distributions:
$$\delta = max |D_{obs}(x) - D_{src}(x)|$$

Now we can define DNA sequence reconstruction problem in a better way:

\begin{definition}{DNA sequence reconstruction problem 2.}
Given set of reads $\mathcal{F}$ and error rate $\varepsilon$, find
a sequence $S$ and $\varepsilon$-valid layout which has minimal
maximum deviation between observed and source distrition of reads.
\end{definition}



\cite{miller2010assembly}

\cite{Medvedev2009}
